Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-08-07 manuscripts/reply/reply-abstract.tex
More editing of abstract and main text of reply ms.

Establishing that a set of population splitting events occurred at the same
time can be a potentially persuasive argument that a common process affected
the populations.
Recently, \citet{Oaks2012} assessed the ability of an approximate-Bayesian
model choice method (\msb) to estimate such a pattern of simultaneous
divergence across taxa, to which \citet{Hickerson2013} responded.
Both papers agree that the primary inference enabled by the method is very
sensitive to prior assumptions and often erroneously supports shared
divergences across taxa when prior uncertainty about divergence times
is represented by a uniform distribution.
However, the papers differ about the best explanation and solution for this
problem.
\citet{Hickerson2013} suggested numerical approximation error causes \msb
analyses to be biased toward models of clustered divergences because the
method's rejection algorithm is unable to adequately sample the parameter space
of richer models within reasonable computational limits when using broad
uniform priors on divergence times (Hypothesis~\ref{hypError}).
As a potential solution, they proposed a model-averaging approach that uses
narrow, empirically informed uniform priors.
\citet{Oaks2012} suggested the method's behavior was caused by the strong
weight of uniformly distributed priors on divergence times leading to smaller
marginal likelihoods (and thus smaller posterior probabilities) of models with
more divergence time parameters (Hypothesis~\ref{hypML}); they proposed
alternative prior probability distributions to avoid such strongly weighted
posteriors.
% \citet{Oaks2012} suggested the method's behavior was due to the sensitivity of
% marginal likelihoods to the uniformly distributed priors on divergence times
% (Hypothesis~\ref{hypML}), and proposed alternative prior probability
% distributions to mitigate this sensitivity.
Here, we use analyses of simulated and empirical data to show that the tendency
of \msb analyses to support models of shared divergences is primarily due to
Hypothesis~\ref{hypML}, whereas Hypothesis~\ref{hypError} is an untenable
explanation for the bias.
I.e., if our prior assumptions place too much weight in unlikely regions of
parameter space such that the true posterior supports the wrong model of
evolutionary history, no amount of computation can rescue our inference. 
We also demonstrate the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} does not mitigate the method's tendency to erroneously
support models of highly clustered divergences.
Furthermore, we show this approach is dangerous in the sense that the
empirically-derived uniform priors often exclude from consideration the true
values of the divergence-time parameters.
Fortunately, as predicted by fundamental principles of Bayesian model choice,
more flexible distributions that accommodate prior uncertainty about parameters
without placing excessive weight in vast regions of parameter space with low
likelihood increase the method's robustness and power to detect temporal
variation in divergences \citep{Oaks2014dpp}.

>>>>

Establishing that a set of population splitting events occurred at the same
time can be a potentially persuasive argument that a common process affected
the populations.
Recently, \citet{Oaks2012} assessed the ability of an approximate-Bayesian
model choice method (\msb) to estimate such a pattern of simultaneous
divergence across taxa, to which \citet{Hickerson2013} responded.
Both papers agree that the primary inference enabled by the method is very
sensitive to prior assumptions and often erroneously supports shared
divergences across taxa when prior uncertainty about divergence times
is represented by a uniform distribution.
However, the papers differ about the best explanation and solution for this
problem.
\citet{Oaks2012} suggested the method's behavior was caused by the strong
weight of uniformly distributed priors on divergence times leading to smaller
marginal likelihoods (and thus smaller posterior probabilities) of models with
more divergence time parameters (Hypothesis~\ref{hypML}); they proposed
alternative prior probability distributions to avoid such strongly weighted
posteriors.
\citet{Hickerson2013} suggested numerical approximation error causes \msb
analyses to be biased toward models of clustered divergences because the
method's rejection algorithm is unable to adequately sample the parameter space
of richer models within reasonable computational limits when using broad
uniform priors on divergence times (Hypothesis~\ref{hypError}).
As a potential solution, they proposed a model-averaging approach that uses
narrow, empirically informed uniform priors.
Here, we use analyses of simulated and empirical data to demonstrate that the
approach of \citet{Hickerson2013} does not mitigate the method's tendency to
erroneously support models of highly clustered divergences, and is dangerous in
the sense that the empirically-derived uniform priors often exclude from
consideration the true values of the divergence-time parameters.
Our results also show that the tendency of \msb analyses to support models of
shared divergences is primarily due to Hypothesis~\ref{hypML}, whereas
Hypothesis~\ref{hypError} is an untenable explanation for the bias.
Overall, this series of papers demonstrate that if our prior assumptions place
too much weight in unlikely regions of parameter space such that the exact
posterior supports the wrong model of evolutionary history, no amount of
computation can rescue our inference. 
Fortunately, as predicted by fundamental principles of Bayesian model choice,
more flexible distributions that accommodate prior uncertainty about parameters
without placing excessive weight in vast regions of parameter space with low
likelihood increase the method's robustness and power to detect temporal
variation in divergences.

### 2014-08-07 manuscripts/reply/reply-main.tex
More editing of abstract and main text of reply ms.

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method is very sensitive to prior assumptions and often
supports shared divergences across taxa that diverged randomly over broad time
periods, to which \citet{Hickerson2013} responded.
\citet{Oaks2012} and \citet{Hickerson2013} agree on the fundamental
methodological point about the model selection performed in \msb:
\begin{itemize}
   \item Representing prior uncertainty about divergence-time parameters with a
       uniform distribution can lead to spurious support for models with few
       divergence events shared across taxa. Thus, the primary inference
       enabled by the approach is very sensitive to the priors on divergence
       times.
\end{itemize}
However, the two papers suggest alternative mechanisms by which the priors on
divergence times cause this behavior:
\begin{enumerate}[label=\textsl{\textbf{Hypothesis \arabic*)}},ref=\arabic*,align=left]
    \item \textsl{\textbf{Numerical approximation error}} \citep{Hickerson2013}:
        Under broad uniform priors, the rejection algorithm implemented in \msb
        is unable to adequately sample the space of the models within
        reasonable computational time, which leads to bias toward models with
        fewer divergence-time parameters because they are better sampled.
        \label{hypError}
    \item \textsl{\textbf{Strongly weighted marginal likelihoods}}
        \citep{Oaks2012}:
        The uniform priors on divergence times lead to very small marginal
        likelihoods (and thus smaller posterior probabilities) of models with
        many divergence-time parameters.  The likelihood of these models is
        ``averaged'' over a much greater parameter space in which there is a
        large amount of prior weight and small probability of producing the
        data.  \citep{Jeffreys1939,Lindley1957}. \label{hypML}
\end{enumerate}
In Hypothesis~\ref{hypError}, the problem is numerical approximation error due
to insufficient computation.
In this scenario, given data from taxa that diverged randomly through time, the
exact (true) posterior supports a model with many divergence-time parameters,
but we are unable to accurately approximate this posterior.
In Hypothesis~\ref{hypML}, the problem is more fundamental; given data from
taxa that diverged randomly through time, the exact posterior supports a model
with simultaneous divergences across taxa.
I.e., when accommodating prior uncertainty about divergence times with a
uniform distribution, the exact posterior from Bayes' rule leads us to the
wrong conclusion about evolutionary history.
Such posterior support for simultaneous divergence, even if ``correct'' from
the perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.

@@

\section{Conclusions}
We demonstrate how the approximate Bayesian model-choice method implemented in
\msb can spuriously support models with less parameter space.
This is caused by the use of uniform priors on divergence times.
Uniform distributions necessitate the use of priors that place high density in
unlikely regions of parameter space, less the risk of excluding the true
divergence times \emph{a priori}.
These broad uniform priors reduce the marginal likelihoods of models with more
divergence-time parameters.
We show that the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} does not mitigate this bias, but rather causes it to
manifest by sampling predominantly from models that often exclude the true
values of the divergence times.
Our results show that it is difficult to choose an uniformly distributed prior
on divergence times that is broad enough to confidently contain the true values
of parameters while being narrow enough to avoid strongly weighted and
misleading posterior support for models with less parameter space.
More generally, it is important to carefully choose prior assumptions about
parameters in Bayesian model selection, because they can strongly influence the
\emph{true} posterior probabilities of the models we seek to estimate.
No amount of computation can rescue our inference if our prior assumptions
place too much weight in unlikely regions of parameter space such that the
exact posterior supports the wrong model of evolutionary history.

>>>>

Recently, \citet{Oaks2012} used this model-choice framework to study 22 pairs
of vertebrate lineages distributed across the Philippines; they also studied
the behavior of the \msb approach using computer simulations.
They found the method is very sensitive to prior assumptions and often
supports shared divergences across taxa that diverged randomly over broad time
periods, to which \citet{Hickerson2013} responded.
\citet{Oaks2012} and \citet{Hickerson2013} agree on the fundamental
methodological point about the model selection performed in \msb:
\begin{itemize}
   \item Representing prior uncertainty about divergence-time parameters with a
       uniform distribution can lead to spurious support for models with few
       divergence events shared across taxa. Thus, the primary inference
       enabled by the approach is very sensitive to the priors on divergence
       times.
\end{itemize}
However, the two papers suggest alternative mechanisms by which the priors on
divergence times cause this behavior:
\begin{enumerate}[label=\textsl{\textbf{Hypothesis \arabic*)}},ref=\arabic*,align=left]
    \item \textsl{\textbf{Strongly weighted marginal likelihoods}}
        \citep{Oaks2012}:
        The uniform priors on divergence times lead to very small marginal
        likelihoods (and thus smaller posterior probabilities) of models with
        many divergence-time parameters.  The likelihood of these models is
        ``averaged'' over a much greater parameter space in which there is a
        large amount of prior weight and small probability of producing the
        data.  \citep{Jeffreys1939,Lindley1957}. \label{hypML}
    \item \textsl{\textbf{Numerical approximation error}} \citep{Hickerson2013}:
        Under broad uniform priors, the rejection algorithm implemented in \msb
        is unable to adequately sample the space of the models within
        reasonable computational time, which leads to bias toward models with
        fewer divergence-time parameters because they are better sampled.
        \label{hypError}
\end{enumerate}
In Hypothesis~\ref{hypError}, the problem is numerical approximation error due
to insufficient computation.
In this scenario, given data from taxa that diverged randomly through time, the
exact (true) posterior supports a model with many divergence-time parameters,
but we are unable to accurately approximate this posterior.
In Hypothesis~\ref{hypML}, the problem is more fundamental; given data from
taxa that diverged randomly through time, the exact posterior supports a model
with simultaneous divergences across taxa.
I.e., when accommodating prior uncertainty about divergence times with a
uniform distribution, the exact posterior from Bayes' rule leads us to the
wrong conclusion about evolutionary history.
Such posterior support for simultaneous divergence, even if ``correct'' from
the perspective of Bayesian model choice, does not provide the biogeographical
insights that a researcher who employs \msb seeks to gain.

@@

\section{Conclusions}
We demonstrate how the approximate Bayesian model-choice method implemented in
\msb can spuriously support models with less parameter space.
This is caused by the use of uniform priors on divergence times.
Uniform distributions necessitate the use of priors that place high density in
unlikely regions of parameter space, less the risk of excluding the true
divergence times \emph{a priori}.
These broad uniform priors reduce the marginal likelihoods of models with more
divergence-time parameters.
We show that the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} does not mitigate this bias, but rather causes it to
manifest by sampling predominantly from models that often exclude the true
values of the divergence times.
Our results show that it is difficult to choose an uniformly distributed prior
on divergence times that is broad enough to confidently contain the true values
of parameters while being narrow enough to avoid strongly weighted and
misleading posterior support for models with less parameter space.
More generally, it is important to carefully choose prior assumptions about
parameters in Bayesian model selection, because they can strongly influence the
posterior probabilities of the models we seek to compare.
No amount of computation can rescue our inference if our prior assumptions
place too much weight in unlikely regions of parameter space such that the
exact posterior supports the wrong model of evolutionary history.
